### 2023-11-20 python_excel_ite_r1.tex
created r1
edits for revision

Spreadsheet based modeling has transformed the teaching of management science in business schools over the past twenty plus years. The late 1990's saw first editions of game changing textbooks \cite{winstonPracticalManagementScience2018,ragsdaleSpreadsheetModelingDecision2017} focused on the use of spreadsheets in teaching modeling and management science.  A series of conferences focused on teaching management science with spreadsheets started at Dartmouth in 1998 and several presentations from that conference made up the bulk of papers in the first two issues of the INFORMS Transactions on Education (ITE) journal. The next few decades saw numerous textbooks \cite{powellBusinessAnalyticsArt2019, camm2020business} and teaching papers in journals such as \href{https://pubsonline.informs.org/journal/ited}{ITE} that prominently featured spreadsheet based modeling.

@@

Numerous R and Python based data science courses have been developed and offered at universities and through online learning platforms such as \href{https://www.coursera.org/browse/data-science}{Coursera}, \href{https://www.edx.org/learn/data-science}{EdX}, \href{https://www.datacamp.com/}{DataCamp}, \href{https://www.udemy.com/topic/data-analysis/}{Udemy} and many others. In 2015, at Oakland University, we created a course called \href{http://www.sba.oakland.edu/faculty/isken/courses/pcda/}{Practical Computing for Data Analytics} (PCDA) that introduced both R and Python (along with a little Linux) for use in data analytics work. This new course was meant as a follow up to the \href{http://www.sba.oakland.edu/faculty/isken/courses/ba/}{spreadsheet based modeling course} we have had in place since 2001.

@@

Python has continued to make inroads to analytical territory that was formerly dominated by spreadsheets. The widely used Python data analysis library, pandas, was created by Wes McKinney while working as a quantitative financial analyst \cite{mckinney2022python}. Python is widely used for data wrangling, visualization and predictive modeling using various flavors of regression, tree based models and neural networks. You can get a very good sense of this by visiting the very popular \href{https://www.reddit.com/r/datascience/}{r/datascience subreddit on Reddit} -- a well known online forum. But, it got me to thinking. What about those things for which Excel is particularly well suited such as building formula based models and doing sensitivity analysis on these models? What would those look like in Python? The importance of modeling, articulated so well in the pages of this journal \cite{powellTeachingModelingManagement2001}, has not abated. What role does Python have to play in supporting the kind of modeling taught for years in our business schools?

@@

Recently there has been more recognition of the need for data scientists to learn basic software engineering principes \cite{nelsonSoftwareEngineeringData2024,treadwaySoftwareEngineeringData2023,rodriguesBuildingReproducibleAnalytical2023}.  The image of newly minted data science graduates developing spaghetti coded Jupyter notebooks not nearly ready for being put into production has reached \href{https://www.reddit.com/r/ProgrammerHumor/comments/y2xe3n/like_every_time_ever_when_the_devops_engineer/}{meme worthy status}. I hoped to address some of these issues in the EwP module. Students would be exposed to some simple software engineering concepts during the process of building and analyzing familiar models. This is really no different than the integration of advanced Excel functions and techniques with spreadsheet modeling tutorials. Just as spreadsheet modeling students were able to easily wow their coworkers and managers with INDEX-MATCH and intermediate Pivot Table skills, I hoped that my students ability to properly document a Python function and use git and GitHub would give them a bit more professional credibility.

@@

At our institution, The EwP module is part of a business course entitled  \href{http://www.sba.oakland.edu/faculty/isken/courses/aap}{Advanced Analytics with Python} (AAP). The students taking this course have already taken the PCDA course within which they learned fundamental Python programming within the context of data analytics over seven weeks. The PCDA course ends with intro to the \href{https://scikit-learn.org/stable/index.html}{scikit-learn library} for doing predictive modeling in Python and the AAP course begins where the PCDA course ends. The EwP module could also be used as part of a semester long Python based analytics course. Most AAP students have also already taken   \href{http://www.sba.oakland.edu/faculty/isken/courses/ba}{Business Analytics} (BA) which is a spreadsheet based introduction to business analytics. The BA and PCDA courses have historically been offered in face to face mode in a computer teaching lab. Since 2020, these two courses have been also offered as online, asynchronous courses. The newer AAP course has been offered since 2021 in an online asynchronous mode. All three courses have extensive course websites that are publicly accessible, including access to all of the video content and supporting files. These courses serve upper level undergraduate and graduate students. Many are from the School of Business Administration but they draw students from many programs across our campus.

@@

\begin{itemize}
	\item variables, arithmetic and boolean operators, basic data types,
	\item data structures such as lists, dictionaries, tuples, NumPy arrays and pandas dataframes,
	\item flow control such as branching with \texttt{if \ldots else} and iterating with \texttt{for} and \texttt{while}
	\item module imports,
	\item using built in functions and accessing methods and properties of objects,
	\item creating functions,
	\item basic use of NumPy and pandas for data wrangling and analysis,
	\item basic plotting with matplotlib.
\end{itemize}

@@

All of the Python modules in the PCDA course as well as the entire AAP course are taught using a combination of Jupyter notebooks, Python scripts, and videos that walk readers through the notebooks. Jupyter notebooks, part of \href{https://jupyter.org/}{Project Jupyter}, are an example of \textit{literate computing} \cite{perez2015project} and allow one to create rich interactive documents that support iterative exploration and development.

@@

The ability to weave standard explanatory language with formal programming languages has made notebook computing a popular tool in the data science education community \cite{perkel2018jupyter}. Jupyter notebooks can also include equations written in \LaTeX, data visualizations, and other multi-media. The technology is free, open source, runs on most computing platforms, and has a huge user community that has created a large number of resources for students, faculty and industry practitioners. There are countless introductory tutorials for Jupyter notebooks, but the official Project Jupyter Documentations \cite{jupyterteamProjectJupyterDocumentation2015} is a very good place to start if you are new to notebook computing. As with all tools, there are pros and cons to their use. See \cite{rule2018ten, barba2019teaching} and \cite{johnson2020benefits} for guidance on Jupyter notebook best practices for both research and teaching.

\begin{figure*}[!htbp]
	\centering
	\includegraphics[scale=0.5]{images/cells_edit_mode.png}
	\caption{Notebook cells in edit model}
	\label{fig:cells_edit_mode}
\end{figure*}

\begin{figure*}[!htbp]
	\centering
	\includegraphics[scale=0.5]{images/cells_after_running.png}
	\caption{Notebook cells after execution}
	\label{fig:cells_after_running}
\end{figure*}

@@

There are two overarching learning objectives for this \href{http://www.sba.oakland.edu/faculty/isken/courses/aap/mod3a_whatif.html}{first EwP submodule}. Students learn how Python might be used to do something they are quite familiar with doing in a spreadsheet -- model building and sensitivity analysis. At the same time, students encounter, as needed, more advanced Python programming concepts and techniques. In this way, the computational problem to be solved drives the introduction of more advanced Python concepts and techniques instead of such things being presented in a vacuum. I make it a point to show that there are often multiple possible approaches to a computational problem by being willing to abandon one approach and trying another. Emphasis is given to the lessons learned from each path explored. Every notebook is accompanied by screencasts that walk the students through the material and act as a guided tour with challenges to overcome along the way. While all of the notebooks are publicly available (see Section \ref{sec:software}), links to an html version of each of the notebooks are provided throughout the document.

@@

The stage is set with an overview of a typical spreadsheet based model that most of the students have seen before. The notebook also includes a screenshot (Figure \ref{fig:xl_model}) of this Excel model.

@@

At this point we discuss a fundamental difference between the computing model of Excel and that of Jupyter notebooks. While spreadsheets respond automatically to changes, Jupyter notebooks require rerunning all code cells that include or follow the changed code. This difference in behavior between Jupyter notebooks and spreadsheets might be one of the most important concepts for students to internalize. A well known challenge, often referred to as the \textit{hidden state problem}, associated with computational notebooks is the ease of running code cells out of order and introducing inadvertent errors \cite{johnson2020benefits,grusDonNotebooks2018}. This point along with strategies to avoid it are discussed during the first week of the course which focuses on \href{http://www.sba.oakland.edu/faculty/isken/courses/aap/jupyter_notebooks.html}{notebook computing best practices as well as a historical look} at computational notebooks.

@@

Since order quantity is the key decision variable, we might want to see how profit changes for different order quantities. In Excel, the Data Table tool provides a familiar and easy way to do this. In a Data Table, a range of order quantities can be used as row or column input and one or more output values can be computed. In Python, the vectorized nature of the \href{https://numpy.org/}{NumPy} library \cite{harris2020array} provides a simple way to accomplish the same thing. A range of order quantities is created as a NumPy array that can be used directly to compute vectors of all of the intermediate and final output variables. Much like Data Tables, vectorized computations avoid having to explicitly iterate, or loop, through a collection of input values.

@@

A few design dilemmas are posed and some non-working approaches illustrated as we try to decide on how to implement an n-way data table function for this new object oriented model. These meta reflections are intentionally included to better mimic actual software development and reinforce the idea that software development is much more than simply writing code to implement a perfectly thought out design. We forge ahead with some preliminary ideas and are confronted with an intermediate problem related to generating a list of dictionaries that represent combinations of inputs, or \textit{scenarios}, to evaluate. This provides a great opportunity to remind students of one of the great strengths of the open source software ecosystem -- we can leverage work done in other packages and even look at actual source code to see how something was implemented. In this case, the well known \href{https://scikit-learn.org/stable/}{scikit-learn} package \cite{scikit-learn} has a \code{ParameterGrid} function that solves our scenario generation problem quite nicely. This example also reminds the students of the value of being able to make sense of API documentation and to understand object oriented code written by others.

@@

Before trying to build our \code{goal\_seek} function we take a brief detour into basic root finding algorithms through packages like \href{https://docs.scipy.org/doc/scipy/tutorial/optimize.html#root-finding}{SciPy} \cite{2020SciPy-NMeth} as well as in \href{https://github.com/patrickwalls/mathematical-python/}{open source notebooks and scripts} that are widely available in the Python data science ecosystem \cite{wallsMathematicalPython2023}. We compare the output of Excel's Goal Seek with different starting values to the values obtained by different root finding algorithms implemented in Python. The importance of starting values and also the range of different root finding algorithms available is often surprising to business students who have spent most of their analytical life in Excel.

@@

In the Excel modeling world, Monte-Carlo simulation can be done without add-ins, but packages like \href{https://lumivero.com/products/at-risk/}{@Risk} can make the process much easier. In particular, random variable generation for a wide range of probability distributions is facilitated by add-ins like @Risk. In Python, the \href{https://numpy.org/doc/stable/reference/random/generator.html}{NumPy package provide random variate generation} for a large number of distributions and similarly, \href{https://docs.scipy.org/doc/scipy/reference/stats.html}{SciPy provides functions for computing distribution related quantities} for many probability distributions. Much like we did when creating the \code{data\_table} function, we start with adding uncertainty to a single input variable in the bookstore model and rely on NumPy's vectorized computing capabilities to generate the profit associated with each realization of the random input. Then we can do standard statistical analysis of the simulation output using Python packages such as pandas and SciPy. In spreadsheet based simulation textbooks, the notion of the ``Flaw of Averages'' popularized by Sam Savage \cite{savageFlawAveragesWhy2012} is often illustrated. We do the same with our Python based model, showing how replacing random demand by its mean can result in a wildly optimistic estimate of the mean profit found through analyzing the simulation output. Much like doing simulation in Excel without add-ins, using Python makes transparent the quite simple and brute force nature of Monte-Carlo simulation.

@@

Now that we have created some useful modeling functionality, our goal is to make it easier for a modeler to actually use the code. Certainly we do not want to rely on copying, pasting and modifying code when building a new model. In the Python world, the natural next step is to deploy our code as a Python \textit{package} \cite{buezenWelcomePythonPackages2022} that can then be imported like any other Python library.

Creating a deployable Python package is a good exercise in designing and creating a reusable software artifact. Thinking about how others will interact with your software usually leads to better designed software.  In \href{http://www.sba.oakland.edu/faculty/isken/courses/aap/mod3b_whatif_packaging.html}{this submodule} we create a Python package that can be imported and used by other modelers. This will require some design changes to our software and will also involve moving code out of Jupyter notebooks and into Python script files. While Jupyter notebooks can be integrated into production environments \cite{uffordInteractiveNotebookInnovation2019}, it is much more common for code to live in Python script files. In addition to creating a deployable package, this submodule also discusses the importance of documentation in its many forms in a Python project.

@@

The first notebook in this submodule starts by explaining what Python packages are and their role in sharing code. We then revisit a concept from earlier in the course -- creating a good project folder structure. For this we use what is known as a \href{https://cookiecutter.readthedocs.io/en/stable/}{cookiecutter} that can automatically generate a project folder structure and key files from the answers to a few prompts \cite{audreyroyandcookiecuttercommunityCookiecutterBetterProject2012}. There are a few different folder layouts used in Python projects and there is a bit of discussion on the folder structure we will use and resources given for exploring this issue further. Students are reminded of the importance of version control and we initialize a git repository for our budding package. The basics of version control were already covered in the first week of the course. My experience is that most students have little exposure to software project management concepts such as project folder structures and version control. This can lead to a jumble of files with incoherent names used as a proxy for version control -- \code{whatif\_v1.py}, \code{whatif\_v2.py}, \code{whatif\_final.py}, \code{whatif\_finalfinal.py}, and so on.

@@

This notebook ends by pointing the students to another notebook in which our newly deployed whatif package is used for a completely different model -- the New Car Simulation that has been a part of Winston and Albright's textbooks for many years \cite{winstonPracticalManagementScience2018,albrightBusinessAnalyticsData2016}. This new model differs structurally from the Bookstore Model, but is representative of many financial models in that cash flows over multiple years need to be computed and summarized. This helps cement the idea that there is value in creating reusable code.

@@

A single Jupyter notebook is used to give a taste of using Python to automate the process of working with Excel files by tackling each of the four examples above. Each of these examples are based on a real problems I encountered either in research or industrial projects. Several web based resources are shared including the \href{https://pbpython.com/}{Practical Business Python blog} written by Chris Moffitt, a data analytics professional \cite{moffittPracticalBusinessPython2022}. This blog, which focuses on helping Excel-centric analytics professionals transition some of their work to Python, includes numerous posts on different aspects of Python and Excel integration.

>>>>

Spreadsheet based modeling has transformed the teaching of management science in business schools over the past twenty plus years. The late 1990's saw first editions of game changing textbooks \citep{winstonPracticalManagementScience2018,ragsdaleSpreadsheetModelingDecision2017} focused on the use of spreadsheets in teaching modeling and management science.  A series of conferences focused on teaching management science with spreadsheets started at Dartmouth in 1998 and several presentations from that conference made up the bulk of papers in the first two issues of the INFORMS Transactions on Education (ITE) journal \citep{baker2000:lp,bell2000:statistics,evans2000:simulation,carraway2000:mba,ragsdale2001:teaching,savage2001:blitzograms}. The next few decades saw numerous textbooks, including {\citet{powellBusinessAnalyticsArt2019, camm2020business}}, and teaching papers in journals such as \href{https://pubsonline.informs.org/journal/ited}{ITE} that prominently featured spreadsheet based modeling.

@@

Numerous R and Python based data science courses have been developed and offered at universities and through online learning platforms such as \href{https://www.coursera.org/browse/data-science}{Coursera}, \href{https://www.edx.org/learn/data-science}{EdX}, \href{https://www.datacamp.com/}{DataCamp}, \href{https://www.udemy.com/topic/data-analysis/}{Udemy} and many others. One particularly relevant example are the courses in computation offered by the Naval Postgraduate School and described in \citet{alderson2022:interactive_computing}. In their courses, Jupyter notebooks (discussed in more detail in the next section) play a prominent role in providing an interactive development and learning environment for Python within an operations research context. 
In 2015, at Oakland University, we created a course called \href{http://www.sba.oakland.edu/faculty/isken/courses/pcda/}{Practical Computing for Data Analytics} (PCDA) that introduced both R and Python (along with a little Linux) for use in data analytics work. This new course was meant as a follow up to the \href{http://www.sba.oakland.edu/faculty/isken/courses/ba/}{spreadsheet based modeling course} we have had in place since 2001.

@@

Python has continued to make inroads to analytical territory that was formerly dominated by spreadsheets. The widely used Python data analysis library, pandas, was created by Wes McKinney while working as a quantitative financial analyst \citep{mckinney2022python}. Python is widely used for data wrangling, visualization and predictive modeling using various flavors of regression, tree based models and neural networks. You can get a very good sense of this by visiting the very popular \href{https://www.reddit.com/r/datascience/}{r/datascience subreddit on Reddit} -- a well known online forum. But, it got me to thinking. What about those things for which Excel is particularly well suited such as building formula based models and doing sensitivity analysis on these models? What would those look like in Python? The importance of modeling, articulated so well by \citet{powellTeachingModelingManagement2001} in the pages of this journal, has not abated. What role does Python have to play in supporting the kind of modeling taught for years in our business schools?

@@

Recently there has been more recognition of the need for data scientists to learn basic software engineering principes \citep{nelsonSoftwareEngineeringData2024,treadwaySoftwareEngineeringData2023,rodriguesBuildingReproducibleAnalytical2023}.  The image of newly minted data science graduates developing spaghetti coded Jupyter notebooks not nearly ready for being put into production has reached \href{https://www.reddit.com/r/ProgrammerHumor/comments/y2xe3n/like_every_time_ever_when_the_devops_engineer/}{meme worthy status}. I hoped to address some of these issues in the EwP module. Students would be exposed to some simple software engineering concepts during the process of building and analyzing familiar models. This is really no different than the integration of advanced Excel functions and techniques with spreadsheet modeling tutorials. Just as spreadsheet modeling students were able to easily wow their coworkers and managers with INDEX-MATCH and intermediate Pivot Table skills, I hoped that my students ability to properly document a Python function and use git and GitHub would give them a bit more professional credibility.

@@

At our institution, The EwP module is part of a business course entitled  \href{http://www.sba.oakland.edu/faculty/isken/courses/aap}{Advanced Analytics with Python} (AAP). The students taking this course have already taken the PCDA course within which they learned fundamental Python programming within the context of data analytics over seven weeks. The PCDA course ends with intro to the \href{https://scikit-learn.org/stable/index.html}{scikit-learn library} for doing predictive modeling in Python and the AAP course begins where the PCDA course ends. The EwP module follows immediately after the modules on machine learning and builds on the Python fundamentals which were covered in the PCDA course. The EwP module could also be used as part of a semester long Python based analytics course in which the Python modules covered in the PCDA course are followed by more advanced topics such as EwP or other topics in our AAP course. The only reason the EwP module is not included in the PCDA course is that Python is only covered in half of that course - the other half being an introduction to R and Linux. If you visit the open access \href{http://www.sba.oakland.edu/faculty/isken/courses/pcda/}{PCDA course website}  and \href{http://www.sba.oakland.edu/faculty/isken/courses/aap/}{AAP course website}, you can see the topic outline for each course and how the EwP module is positioned.

Most AAP students have also already taken   \href{http://www.sba.oakland.edu/faculty/isken/courses/ba}{Business Analytics} (BA) which is a spreadsheet based introduction to business analytics (\cite{isken2003:muddy,isken2014:lab}). The BA and PCDA courses have historically been offered in face to face mode in a computer teaching lab. Since 2020, these two courses have been also offered as online, asynchronous courses. The newer AAP course has been offered since 2021 in an online asynchronous mode. All three courses have extensive course websites that are publicly accessible, including access to all of the video content and supporting files. These courses serve upper level undergraduate and graduate students. Many are from the School of Business Administration but they draw students from many programs across our campus.

@@

\begin{itemize}
	\item familiarity with Jupyter notebooks and an integrated development environment such as \href{https://www.spyder-ide.org/}{Spyder}, \href{https://code.visualstudio.com/}{VSCode} or \href{https://www.jetbrains.com/pycharm/}{PyCharm},
	\item variables, arithmetic and boolean operators, basic data types,
	\item data structures such as lists, dictionaries, tuples, NumPy arrays and pandas dataframes,
	\item flow control such as branching with \texttt{if \ldots else} and iterating with \texttt{for} and \texttt{while}
	\item module imports,
	\item using built in functions and accessing methods and properties of objects,
	\item creating functions,
	\item basic use of NumPy and pandas for data wrangling and analysis,
	\item basic plotting with matplotlib.
\end{itemize}

@@

All of the Python modules in the PCDA course as well as the entire AAP course are taught using a combination of Jupyter notebooks, Python scripts, and videos that walk readers through the notebooks. Jupyter notebooks, part of \href{https://jupyter.org/}{Project Jupyter}, are an example of \textit{literate computing} (\cite{perez2015project}) and allow one to create rich interactive documents that support iterative exploration and development.

@@

The ability to weave standard explanatory language with formal programming languages has made notebook computing a popular tool in the data science education community (\cite{perkel2018jupyter}). Jupyter notebooks can also include equations written in \LaTeX, data visualizations, and other multi-media. The technology is free, open source, runs on most computing platforms, and has a huge user community that has created a large number of resources for students, faculty and industry practitioners. There are countless introductory tutorials for Jupyter notebooks, but the official Project Jupyter Documentation (\cite{jupyterteamProjectJupyterDocumentation2015}) is a very good place to start if you are new to notebook computing. As with all tools, there are pros and cons to their use. See \cite{rule2018ten, barba2019teaching} and \cite{johnson2020benefits} for guidance on Jupyter notebook best practices for both research and teaching.

\begin{figure*}[!htbp]
\centering
\includegraphics[scale=0.5]{cells_edit_mode}
\caption{Notebook cells in edit model}
\label{fig:cells_edit_mode}
\end{figure*}

\begin{figure*}[!htbp]
\centering
\includegraphics[scale=0.5]{cells_after_running}
\caption{Notebook cells after execution}
\label{fig:cells_after_running}
\end{figure*}

@@

There are two overarching learning objectives for this \href{http://www.sba.oakland.edu/faculty/isken/courses/aap/mod3a_whatif.html}{first EwP submodule} - see Figure \ref{fig:objectives_whatif} for how the objectives are presented to students via the course website. Students learn how Python might be used to do something they are quite familiar with doing in a spreadsheet -- model building and sensitivity analysis. At the same time, students encounter, as needed, more advanced Python programming concepts and techniques. In this way, the computational problem to be solved drives the introduction of more advanced Python concepts and techniques instead of such things being presented in a vacuum. I make it a point to show that there are often multiple possible approaches to a computational problem by being willing to abandon one approach and trying another. Emphasis is given to the lessons learned from each path explored. Every notebook is accompanied by screencasts that walk the students through the material and act as a guided tour with challenges to overcome along the way. While all of the notebooks are publicly available (see Section \ref{sec:software}), links to an html version of each of the notebooks are provided throughout the document.

\begin{figure}[!htbp]
	\includegraphics[width=1.0\textwidth]{objectives_whatif}
	\caption{Learning objectives for Submodule 1}
	\label{fig:objectives_whatif}
\end{figure}

@@

The stage is set with an overview of a typical spreadsheet based model that most of the students have seen before. The notebook also includes a screenshot (Figure \ref{fig:xl_model}) of this Excel model. Readers are encouraged to click the \href{http://www.sba.oakland.edu/faculty/isken/excel_with_python/what_if_1_model_datatable.html}{html version} link so that you can see all of the notebook details as you are reading through this section.

@@

At this point we discuss a fundamental difference between the computing model of Excel and that of Jupyter notebooks. While spreadsheets respond automatically to changes, Jupyter notebooks require rerunning all code cells that include or follow the changed code. This difference in behavior between Jupyter notebooks and spreadsheets might be one of the most important concepts for students to internalize. A well known challenge, often referred to as the \textit{hidden state problem}, associated with computational notebooks is the ease of running code cells out of order and introducing inadvertent errors -- see \cite{johnson2020benefits,grusDonNotebooks2018}. This point along with strategies to avoid it are discussed during the first week of the course which focuses on \href{http://www.sba.oakland.edu/faculty/isken/courses/aap/jupyter_notebooks.html}{notebook computing best practices as well as a historical look} at computational notebooks.

@@

Since order quantity is the key decision variable, we might want to see how profit changes for different order quantities. In Excel, the Data Table tool provides a familiar and easy way to do this. In a Data Table, a range of order quantities can be used as row or column input and one or more output values can be computed. In Python, the vectorized nature of the \href{https://numpy.org/}{NumPy} library (\cite{harris2020array}) provides a simple way to accomplish the same thing. A range of order quantities is created as a NumPy array that can be used directly to compute vectors of all of the intermediate and final output variables. Much like Data Tables, vectorized computations avoid having to explicitly iterate, or loop, through a collection of input values.

@@

A few design dilemmas are posed and some non-working approaches illustrated as we try to decide on how to implement an n-way data table function for this new object oriented model. These meta reflections are intentionally included to better mimic actual software development and reinforce the idea that software development is much more than simply writing code to implement a perfectly thought out design. We forge ahead with some preliminary ideas and are confronted with an intermediate problem related to generating a list of dictionaries that represent combinations of inputs, or \textit{scenarios}, to evaluate. This provides a great opportunity to remind students of one of the great strengths of the open source software ecosystem -- we can leverage work done in other packages and even look at actual source code to see how something was implemented. In this case, the well known \href{https://scikit-learn.org/stable/}{scikit-learn} package (\cite{scikit-learn}) has a \code{ParameterGrid} function that solves our scenario generation problem quite nicely. This example also reminds the students of the value of being able to make sense of API documentation and to understand object oriented code written by others.

@@

Before trying to build our \code{goal\_seek} function we take a brief detour into basic root finding algorithms through packages like \href{https://docs.scipy.org/doc/scipy/tutorial/optimize.html#root-finding}{SciPy} (\cite{2020SciPy-NMeth}) as well as in \href{https://github.com/patrickwalls/mathematical-python/}{open source notebooks and scripts} that are widely available in the Python data science ecosystem (\cite{wallsMathematicalPython2023}). We compare the output of Excel's Goal Seek with different starting values to the values obtained by different root finding algorithms implemented in Python. The importance of starting values and also the range of different root finding algorithms available is often surprising to business students who have spent most of their analytical life in Excel.

@@

In the Excel modeling world, Monte-Carlo simulation can be done without add-ins, but packages like \href{https://lumivero.com/products/at-risk/}{@Risk} can make the process much easier. In particular, random variable generation for a wide range of probability distributions is facilitated by add-ins like @Risk. In Python, the \href{https://numpy.org/doc/stable/reference/random/generator.html}{NumPy package provide random variate generation} for a large number of distributions and similarly, \href{https://docs.scipy.org/doc/scipy/reference/stats.html}{SciPy provides functions for computing distribution related quantities} for many probability distributions. Much like we did when creating the \code{data\_table} function, we start with adding uncertainty to a single input variable in the bookstore model and rely on NumPy's vectorized computing capabilities to generate the profit associated with each realization of the random input. Then we can do standard statistical analysis of the simulation output using Python packages such as pandas and SciPy. In spreadsheet based simulation textbooks, the notion of the ``Flaw of Averages'' popularized by Sam Savage (\cite{savageFlawAveragesWhy2012}) is often illustrated. We do the same with our Python based model, showing how replacing random demand by its mean can result in a wildly optimistic estimate of the mean profit found through analyzing the simulation output. Much like doing simulation in Excel without add-ins, using Python makes transparent the quite simple and brute force nature of Monte-Carlo simulation.

@@

Now that we have created some useful modeling functionality, our goal is to make it easier for a modeler to actually use the code. Certainly we do not want to rely on copying, pasting and modifying code when building a new model. In the Python world, the natural next step is to deploy our code as a Python \textit{package} (\cite{buezenWelcomePythonPackages2022}) that can then be imported like any other Python library. Figure \ref{fig:objectives_packaging} shows the learning objectives from this \href{http://www.sba.oakland.edu/faculty/isken/courses/aap/mod3b_whatif_packaging.html}{submodule page on the course website}.

\begin{figure}[!htbp]
	\centering
	\includegraphics[scale=0.5]{objectives_packaging}
	\caption{Learning objectives for Submodule 2}
	\label{fig:objectives_packaging}
\end{figure}

Creating a deployable Python package is a good exercise in designing and creating a reusable software artifact. Thinking about how others will interact with your software usually leads to better designed software.  In this submodule we create a Python package that can be imported and used by other modelers. This will require some design changes to our software and will also involve moving code out of Jupyter notebooks and into Python script files. While Jupyter notebooks can be integrated into production environments \cite{uffordInteractiveNotebookInnovation2019}, it is much more common for code to live in Python script files. In addition to creating a deployable package, this submodule also discusses the importance of documentation in its many forms in a Python project.

@@

The first notebook in this submodule starts by explaining what Python packages are and their role in sharing code. We then revisit a concept from earlier in the course -- creating a good project folder structure. For this we use what is known as a \href{https://cookiecutter.readthedocs.io/en/stable/}{cookiecutter} that can automatically generate a project folder structure and key files from the answers to a few prompts (\cite{audreyroyandcookiecuttercommunityCookiecutterBetterProject2012}). There are a few different folder layouts used in Python projects and there is a bit of discussion on the folder structure we will use and resources given for exploring this issue further. Students are reminded of the importance of version control and we initialize a git repository for our budding package. The basics of version control were already covered in the first week of the course. My experience is that most students have little exposure to software project management concepts such as project folder structures and version control. This can lead to a jumble of files with incoherent names used as a proxy for version control -- \code{whatif\_v1.py}, \code{whatif\_v2.py}, \code{whatif\_final.py}, \code{whatif\_finalfinal.py}, and so on.

@@

This notebook ends by pointing the students to another notebook in which our newly deployed whatif package is used for a completely different model -- the New Car Simulation that has been a part of Winston and Albright's textbooks for many years (\cite{winstonPracticalManagementScience2018,albrightBusinessAnalyticsData2016}). This new model differs structurally from the Bookstore Model, but is representative of many financial models in that cash flows over multiple years need to be computed and summarized. This helps cement the idea that there is value in creating reusable code.

@@

A single Jupyter notebook is used to give a taste of using Python to automate the process of working with Excel files by tackling each of the four examples above. Each of these examples are based on a real problems I encountered either in research or industrial projects. Several web based resources are shared including the \href{https://pbpython.com/}{Practical Business Python blog} written by Chris Moffitt, a data analytics professional (\cite{moffittPracticalBusinessPython2022}). This blog, which focuses on helping Excel-centric analytics professionals transition some of their work to Python, includes numerous posts on different aspects of Python and Excel integration.
